---
title: "SOK-1005-Assignment1"
author: "30 | Casper Andresen"
format: pdf
editor: visual
---

```{r}
rm(list=ls())

library(tidyverse)
library(lubridate)
```

Importing the data

```{r}
df_lt <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/tlt/uahncdc_lt_6.0.txt")
df_mt <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/tmt/uahncdc_mt_6.0.txt")
df_t <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/ttp/uahncdc_tp_6.0.txt")
df_ls <- read_table("http://vortex.nsstc.uah.edu/data/msu/v6.0/tls/uahncdc_ls_6.0.txt")

```

Removing unwanted bottom text

```{r}
df_lt <- df_lt[1:which(df_lt$Year %in% "Year")-1, ]
df_mt <- df_mt[1:which(df_mt$Year %in% "Year")-1, ]
df_t <- df_t[1:which(df_t$Year %in% "Year")-1, ]
df_ls <- df_ls[1:which(df_ls$Year %in% "Year")-1, ]
```

Combinding the globe column in each dataframe into a new dataframe and changing them to numeric

```{r}
df_global <- cbind(df_lt$Globe, df_mt$Globe, df_t$Globe, df_ls$Globe)
df_global <- as.data.frame(df_global)

df_global <- df_global %>%
  rename(lt = V1) %>%
  rename(mt = V2) %>%
  rename(t = V3) %>%
  rename(ls = V4)

df_global <- df_global %>%
  mutate_at(vars(lt, mt, t, ls), ~as.numeric(.))
```

Creating a Data variable using the year and month variable from the df_lt dataframe, then move it to the front

```{r}
df_global <- df_global %>% 
  select(lt, mt, t, ls) %>%
  mutate(Date = ymd(paste(df_lt$Year, df_lt$Mo, 1, sep="-"))) %>%
  relocate(Date)

```

Calculating the rolling mean, filter out the dates I want

```{r}
df_global <- df_global %>%
  mutate(average_lt = zoo::rollmean(lt, 13, na.pad = TRUE, align = "right")) %>%
  mutate(average_mt = zoo::rollmean(mt, 13, na.pad = TRUE, align = "right")) %>%
  mutate(average_t = zoo::rollmean(t, 13, na.pad = TRUE, align = "right")) %>%
  mutate(average_ls = zoo::rollmean(ls, 13, na.pad = TRUE, align = "right")) %>%
  filter(between(Date, as.Date("1980-1-1"), as.Date("2022-12-1")))
```

Putting it in a tall format so I can plot them all together

```{r}
df_tall <- df_global %>%
  select(Date, average_lt, average_mt, average_t, average_ls) %>%
  pivot_longer(-Date, names_to = "Location", values_to = "Temp")
```

Drawing the figure

```{r}

df_tall %>%
  ggplot(aes(x=Date, y=Temp, color=Location)) + 
  geom_line() +
  labs(title = "Average temperature over time", subtitle = "Rolling average in the lower layers of the atmosphere. 1980 - 2022") +
  ylab("Global Temperature") + xlab("Year") +
  scale_color_discrete(labels=c("Lower Troposphere", "Mid-Troposphere", "Tropopause", "Lower Stratosphere")) +
  geom_smooth(method = lm, se = FALSE)
```
